Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-05-31 sports_analytics.Rmd
started on win prob and team strength

```{r pkg}
knitr::write_bib(
  c(
    "Lahman", "pitchRx", "retrosheet",
    "teamcolors",
    "ggplot2"
  ),
  file = "pkgs.bib"
)
```

@@

Early work on this topic can be found in @lindsey1963investigation, which used play-by-play data to compute an empirical estimate for the mean number of runs scored in the remainder of the inning for each of these 24 possible states of an inning.  This line of work led to work analyzing all types of common baseball strategies.  For example, many baseball teams elect to attempt a sacrifice bunt with a runner on first and no one out in the inning, with the goal of moving the runner to second base, at the cost of the batter being out.  Theirs, and many subsequent analyses [@tango2007book], conclude that the sacrifice bunt is rarely worth it, because most teams would be expected to score more runs with a runner on first and no outs than they would with a runner on second and one out.

@@

@CarrollPalmer1998 The hidden game of football.  Mentioned by Pattani and popularizing the idea of EPA.

@nflwar, which uses an expected points model as part of the nfl war framework, uses down (1-4, seconds remaining in half, yards from endzone, yards to go (log), indicator for goal to go (GTG), and indicator for under 2 minutes (UTM).

@@

## Win probability

A related, but different, concept is the notion of *win probability*, which is simply an estimate of the probability that either team will win the game, given its current state.

>>>>

```{r pkg}
knitr::write_bib(
  c(
    "Lahman", "pitchRx", "retrosheet",
    "teamcolors",
    "ggplot2",
    "BradleyTerry2"
  ),
  file = "pkgs.bib"
)
```

@@

Early work on this topic can be found in @lindsey1963investigation, which used play-by-play data to compute an empirical estimate for the mean number of runs scored in the remainder of the inning for each of these 24 possible states of an inning.  This line of work led to work analyzing all types of common baseball strategies.  For example, many baseball teams elect to attempt a sacrifice bunt with a runner on first and no one out in the inning, with the goal of moving the runner to second base, at the cost of the batter being out.  Theirs, and many subsequent analyses [@Tango2007book], conclude that the sacrifice bunt is rarely worth it, because most teams would be expected to score more runs with a runner on first and no outs than they would with a runner on second and one out.

@@

@CarrollPalmer1988 The hidden game of football.  Mentioned by Pattani and popularizing the idea of EPA.

@Yurko2019, which uses an expected points model as part of the nfl war framework, uses down (1-4, seconds remaining in half, yards from endzone, yards to go (log), indicator for goal to go (GTG), and indicator for under 2 minutes (UTM).

@@

@fernandez2021epvsoccer A framework for the fine-grained evaluation of the instantaneous expected value of soccer possessions

@decroos2019playeraction Actions Speak Louder than Goals: Valuing Player Actions in Soccer

# Win probability

A related, but different, concept is the notion of *win probability*, which is simply an estimate of the probability that either team will win the game, given its current state. Win probability provides a measure of the impact of sports plays and individual player contributions, as discussed in @Albert2015. Given its popularity, recent books on sports analytics often offers multiple chapters entirely on win probability. These include @Albert2001curve, @Schwarz2005numbers, @Tango2007book, @Albert2017handbook, and @Winston2022. In this section, we discuss notable previous work on win probability in baseball, football, basketball, and several other sports.

## Baseball

@Albert2015 provided a historical overview of the use of win probability in baseball. This notion goes back to at least as early as @Lindsey1961, where the expected win probability after each inning was calculated based on the distribution of runs scored in each inning. @Mills1970 utilized win probability to introduce Player Win Average (PWA), a measure of a player's contribute to the game outcome. In particular, PWA was computed as 
$$
PWA = \frac{Win \ Points}{Win \ Points + Loss \ Points},
$$
where the win and loss points represent how much the player positively and negatively impact their team's probability of winning after each play.

## Football

In football,

Burke WPA

http://www.advancedfootballanalytics.com/2010/01/win-probability-added-wpa-explained.html

@Lock2014 used random forests as a means for estimating WP before each play in a football game.

@Yurko2019

@Yam2019 causal

nflscrapr, nflfastR, nflverse

@Yurko2020 going deep

## Basketball

@Stern1994 brownian

@Deshpande2016 bayesian

@McFarlane2019 nba logistic

## Other sports

https://arxiv.org/pdf/2204.11777.pdf

@Brenzel2019 curling markov

@Maymin2020 esport!

# Team strength

A popular method for estimating team strength in sports is through pairwise evaluations. Perhaps the most famous probability model for predicting the outcome of a paired comparison is the Bradley-Terry model (BTM) [@Bradley1952]. For a pair of players (or teams) $i$ and $j$, let $\Pi_{ij}$ denote the probability that $i$ is preferred to $j$. Then the BTM is a logistic regression model with parameters $\{\beta_i\}$ such that

$$
log\left(\frac{\Pi_{ij}}{\Pi_{ji}}\right) = \beta_i - \beta_j
$$

Here, $exp(\beta_i)$ is often viewed as a representation of player $i$'s ability. The BTM can be implemented in `R` [@Rcitation2022] via the \pkg{BradleyTerry2} package [@R-BradleyTerry2].

@Elo1978

@Glickman1998 state-space model

@Lopez2018 extension to @Glickman1998's framework
